Von Neumann-Morgenstern and Savage Theorems for Causal Decision Making

### 2020-09-18 Artificial_Intelligence_Journal/Draft/AIJ_Draft.tex
version final

\address[1]{Coordinaci\'on de Ciencias Computacionales, Instituto Nacional de Astrof\'isica \'Optica y Electr\'onica ,Luis Enrique Erro 1, Santa Maria Tonanzintla, M\'exico.}

@@

Notice that since the graph is finite, there exist some nodes that do not have causes, which are called \textit{exogenous}. If an event $A$ is caused by some other event, then we say it is \textit{endogenous} and we denote the set of its causes as $Pa(A)$. It is proven in \cite{kiiveri1984recursive} that at least one exogenous node exists in a causal graph.

@@

Also, we consider an extra pair of conditions. The first one, known as Causal Sufficiency, is about the \textit{nature} of the model: for any variable $X$ in the model $\mathcal{G}$ there are no causes of $X$ \textit{outside} of the model $\mathcal{G}$\citep{spirtes2000causation,pearl2009causality,sucar2015probabilistic}. The second required condition is that the causal mechanism remains unchanged after interventions, this is called in \citep{woodward2005making} as \textit{invariantness}, this basically means that an intervention does not \textit{break} the causal mechanism when intervening upon it. These conditions are required in an axiomatic fashion so they will be taken as they are without questioning.

\subsection{Causal Graphical Models}
A Causal Graphical Model (CGM) \citep{koller2009probabilistic,sucar2015probabilistic} consists of a set of random variables $\mathcal{X}=\{ X_1,...,X_n \}$, and a Directed Acyclic Graph (DAG) whose nodes are in correspondence with the variables in $\mathcal{X}$ and whose edges represent relations of cause-effect. Also, the model, which up to this point is nothing more than a Bayesian Network with causal semantics, is enriched with an operator named $do()$ which is defined over graphs and whose action is described as follows: given $\mathbf{X} \subseteq \mathcal{X}$ and $\mathbf{x} = \{ x_{i_1}, x_{i_2}, ... , x_{i_j} \} \in Val(\mathcal{X})$ the action $do(\mathbf{X} = \mathbf{x} )$ corresponds to assign to each $X_j \in \mathbf{X}$ the value $x_{x_{i_j}}$ and to delete every incoming edges into the node corresponding to each $X_j$ in the graph $\mathcal{G}$ \citep{pearl2009causality,koller2009probabilistic,sucar2015probabilistic}. To apply the $do()$ operator over a variable (or set of variables) is also called as an \textit{intervention} over the variable. It is this interventional operator which separates associative models from causal models, since it provides the solution of the “probability raising” attempts to define causality, since if $X$ is a cause for $Y$ then $P(Y | do(X)) > P(Y)$ \citep{pearl2018why}. In fact, using the \textit{Do} operator we give a definition of a variable $X$ causing $Y$.
\begin{definition}
Let $\mathcal{G}$ a Causal Graphical Model and $P_\mathcal{G}$ its corresponding distribution and let $X$ and $Y$ variables in the model. We say that $X$ \textit{causes} $Y$ if 
\[ P_\mathcal{G}(Y | do(X=x)) \neq P_\mathcal{G}(Y | do(X=x')). \]
\end{definition}
It is required that the probability distribution that results from an intervention over a variable is Markov compatible with the graph; this is, the resulting interventional distribution is equivalent to the product of the conditional probability of every variable given its parents in the intervened graph \citep{sucar2015probabilistic}.
\subsubsection{The identifiability problem}
Under what conditions can causal inquiries be answered in terms of purely \textit{observational} data? It is known that if the Markov Causal condition and Causal Faithfulness hold, then it is possible to identify the causal graph up to related variables without direction (Markov equivalence) \citep{peters2012identifiability,mooij2016distinguishing}.

\subsubsection{Do-Calculus}
The Do-Calculus is a set of rules for manipulating probabilistic statements that involve interventions and, under certain conditions, allow them to be transformed into statements that do not involve interventional data \citep{pearl1995causal,pearl2009causality}.

It is important to introduce some notation. Consider a causal graphical model $\mathcal{G}$ and $X,Y,Z$ disjoint sets of nodes of $\mathcal{G}$. We denote by $\mathcal{G}_{\bar{X}}$ the graph that is obtained by deleting from $\mathcal{G}$ all of the edges that enter nodes in $X$. In the same way, $\mathcal{G}_{\underline{X}}$ is the graph obtained by deleting the edges that emerge from $X$. Finally, $\mathcal{G}_{\underline{Z}\bar{X}}$ is the graph obtained by deleting edges incoming into $X$ and outgoing from $Z$.

\begin{theorem}[Do-Calculus, \cite{pearl2009causality}]{\label{docalculus}}
Let $\mathcal{G}$ a CGM and $P_{\mathcal{G}}$ the probability measure induced by the model; then, for disjoint sets of nodes $X,Y,Z,W$ it holds:
\begin{itemize}
\item If for the graph $\mathcal{G}_{\bar{X}}$ it holds that $Y$ is conditionally independent from  $Z$ given $X$ and $W$, then
\[ P_{\mathcal{G}}(Y=y | do(X=x), Z=z, W=w) = P_{\mathcal{G}}(Y=y | do(X=x), W=w). \]
\item If for the graph $\mathcal{G}_{\bar{X}\underline{Z}}$ it holds that $Y$ is conditionally independent from $Z$ given $X$ and $W$, then
\[ P(Y=y | do(X=x), do(Z=z), W=w) = P(Y=y | do(X=x), Z = z, W=w). \]
\item Let $Z(W)$ the set of nodes in  $Z$ that aren't ancestors of any node in  $W$ in the graph $\mathcal{G}_{\bar{X}}$.  If $Y$ is conditionally independent from $Z$ given $X$ and $W$ in the graph $G_{\bar{X}, \bar{Z(W)}}$, then
\[ P(Y=y | do(X=x), do(Z=z), W=w) = P(Y=y | do(X=x), W=w). \]
\end{itemize}
\end{theorem}
\begin{proof}
See \cite{pearl2009causality}.
\end{proof}
\begin{theorem}[Properties of Do-calculus]{\cite{peters2017elements}}\\
The following statements hold:
\begin{itemize}
\item The Do-Calculus is complete; this is, sufficient for deriving every identifiable interventional distribution \citep{huang2006pearl,shpitser2006identification}.
\item There exists an algorithm capable of finding all of the identifiable interventions \citep{tian2002,huang2006pearl}.
\item A necessary and sufficient criteria exists for the identifiability of interventional distributions \citep{shpitser2006identification,huang2006pearl}.
\end{itemize}
\end{theorem}
\begin{proof}
See \cite{peters2017elements}.
\end{proof}
The Do-Calculus rules provide a solution for the identifiability problem:
\begin{corollary} \cite{pearl2009causality}.\\
A distribution $q=P(y_1,...,y_k | do(x_1),...,do(x_n))$ is identifiable in a causal graphical model $\mathcal{G}$ if there exists a finite sequence of transformations, where each one of them corresponds to any of the rules of the Do-Calculus, that reduce $q$ to a purely observational expression.
\end{corollary}

@@

\begin{definition}
We define a Causal Decision Problem (CDP) as a tuple $(\Omega, \mathcal{A}, \mathcal{G},\mathcal{E},\mathcal{C},\succeq)$ where $(\Omega, \mathcal{A}, \mathcal{G},\mathcal{E},\mathcal{C})$ is a Causal Environment and $\succeq$ is a preference relation. 
\end{definition}

@@

\subsection{Causal Reinforcement Learning and Bandit Problems}
It is known that human beings conceive their actions on the world as \textit{intervening} in the world \citep{hagmayer2009decision}. Following this idea, \cite{lattimoreNIPS2016} consider decision problems in which the action to be chosen is an intervention over a known causal graphical model. The agent must choose the intervention that maximizes the value of a \textit{target variable} after a series of learning rounds. They model their problem by considering that choosing an intervention is choosing an arm of a \textit{slot machine}, usually known as Bandit Problems, in which a gambler chooses an arm and gets some reward. From the rewards they estimate probabilities and output an optimal action in the sense of obtaining minimal regret. Their work considers that the causal model is fully known, restriction which is partially relaxed by \cite{sen2017identifying}, and mention that the case in which the causal model is unknown is left as an open question, and this is precisely what we are proposing to answer.

@@

Consider a Causal Model $G$ and its associated distribution $P_G$ and let $C$ be the set of consequences of interest for a decision maker. Then,
\begin{theorem}[Causal von Neumann-Morgenstern Theorem, \cite{pearl2009causality}]{\label{causal_ut}}
If a rational decision maker faces a Causal Environment and if the causal model is known, then the preference relation $\succeq$ satisfies the von Neumann-Morgenstern rationality axioms if and only if:
\begin{eqnarray*}
&a \succeq b&\\
& \textrm{ if and only if }&\\
 &\sum_{c \in C} P(c | do(a))u(c) \geq \sum_{c \in C} P(c | do(b))u(c).&
\end{eqnarray*} 
\end{theorem}
Equivalently, the action that must be chosen is 
\begin{equation}
a^\ast = \textrm{argmax}_{a \in \mathcal{A}} \sum_{c \in C} P(c | do(a))u(c).
\end{equation}
\begin{proof}
This Theorem assumes the existence of a Causal Graphical Model known by the decision maker; then, by Theorem \ref{vNM}, we know that the action that must be chosen is the one that maximizes the expected utility; we must now ask the expected utility of what? The answer lies on the Causal Graphical Model which controls the relation between actions and consequences; if this is the case, then the probability of a consequence given an action $a$ is in fact $P(c|do(a))$. The rest follows by Theorem \ref{vNM}. Details in Section 4.1 of \cite{pearl2009causality}.
\end{proof}

@@

\section{Future Work}
We have left out the implementation of the algorithm stated in Section \ref{random_graphs} we considered the problem of a decision maker situated in a causal environment who does not fully knows the causal model which controls her environment; previous works have considered different levels of uncertainty over the causal model: knowing conditional distributions, knowing a part of the model, knowing the structure of the model or the variables involved. We leave as future work the case in which the variables of the model are not known and we consider that this scenario could be attacked using tools from Bayesian Nonparametrics \citep{hjort2010bayesian,phadia2013prior} in the form of a prior over causal models. If only a part of the set of variables is known, then a Bayesian Nonparametric prior, which generates causal graphs of distinct sizes and complexity, could capture the patterns of causal interaction between known and possible but unknown variables.

>>>>

\address[1]{Coordinaci\'on de Ciencias Computacionales, Instituto Nacional de Astrof\'isica \'Optica y Electr\'onica, Luis Enrique Erro 1, Santa Maria Tonanzintla, M\'exico.}

@@

Notice that since the graph is finite, there exist some nodes that do not have causes, which are called \textit{exogenous}. If an event $A$ is caused by some other event, then we say it is \textit{endogenous} and we denote the set of its causes as $Pa(A)$. It is proven by \cite{kiiveri1984recursive} that at least one exogenous node exists in a causal graph.

@@

Also, we consider an extra pair of conditions. The first one, known as Causal Sufficiency, is about the \textit{nature} of the model: for any variable $X$ in the model $\mathcal{G}$ there are no causes of $X$ \textit{outside} of the model $\mathcal{G}$ \citep{spirtes2000causation,pearl2009causality,sucar2015probabilistic}. The second required condition is that the causal mechanism remains unchanged after interventions, this is called in \citep{woodward2005making} as \textit{invariantness}, this basically means that an intervention does not \textit{break} the causal mechanism when intervening upon it. These conditions are required in an axiomatic fashion so they will be taken as they are without questioning.

\subsection{Causal Graphical Models}
A Causal Graphical Model (CGM) \citep{koller2009probabilistic,sucar2015probabilistic} consists of a set of random variables $\mathcal{X}=\{ X_1,...,X_n \}$, and a Directed Acyclic Graph (DAG) whose nodes are in correspondence with the variables in $\mathcal{X}$ and whose edges represent relations of cause-effect. Also, the model, which up to this point is nothing more than a Bayesian Network with causal semantics, is enriched with an operator named $do()$ which is defined over graphs and whose action is described as follows: given $\mathbf{X} \subseteq \mathcal{X}$ and $\mathbf{x} = \{ x_{i_1}, x_{i_2}, ... , x_{i_j} \} \in Val(\mathcal{X})$ the action $do(\mathbf{X} = \mathbf{x} )$ corresponds to assign to each $X_j \in \mathbf{X}$ the value $x_{x_{i_j}}$ and to delete every incoming edges into the node corresponding to each $X_j$ in the graph $\mathcal{G}$ \citep{pearl2009causality,koller2009probabilistic,sucar2015probabilistic}. To apply the $do()$ operator over a variable (or set of variables) is also called as an \textit{intervention} over the variable. It is this interventional operator which separates associative models from causal models, since it provides the solution of the “probability raising” attempts to define causality, since if $X$ is a cause for $Y$ then $P(Y | do(X)) > P(Y)$ \citep{pearl2018why}. In fact, using the \textit{Do} operator we give a definition of a variable $X$ causing $Y$.
\begin{definition}
Let $\mathcal{G}$ a Causal Graphical Model and $P_\mathcal{G}$ its corresponding distribution and let $X$ and $Y$ variables in the model. We say that $X$ \textit{causes} $Y$ if 
\[ P_\mathcal{G}(Y | do(X=x)) \neq P_\mathcal{G}(Y | do(X=x')). \]
\end{definition}
It is required that the probability distribution that results from an intervention over a variable is Markov compatible with the graph; this is, the resulting interventional distribution is equivalent to the product of the conditional probability of every variable given its parents in the intervened graph \citep{sucar2015probabilistic}.
%\subsubsection{The identifiability problem}
%Under what conditions can causal inquiries be answered in terms of purely \textit{observational} data? It is known that if the Markov Causal condition and Causal Faithfulness hold, then it is possible to identify the causal graph up to related variables without direction (Markov equivalence) \citep{peters2012identifiability,mooij2016distinguishing}.
%
%\subsubsection{Do-Calculus}
%The Do-Calculus is a set of rules for manipulating probabilistic statements that involve interventions and, under certain conditions, allow them to be transformed into statements that do not involve interventional data \citep{pearl1995causal,pearl2009causality}. 
%	
%It is important to introduce some notation. Consider a causal graphical model $\mathcal{G}$ and $X,Y,Z$ disjoint sets of nodes of $\mathcal{G}$. We denote by $\mathcal{G}_{\bar{X}}$ the graph that is obtained by deleting from $\mathcal{G}$ all of the edges that enter nodes in $X$. In the same way, $\mathcal{G}_{\underline{X}}$ is the graph obtained by deleting the edges that emerge from $X$. Finally, $\mathcal{G}_{\underline{Z}\bar{X}}$ is the graph obtained by deleting edges incoming into $X$ and outgoing from $Z$.
%	
%\begin{theorem}[Do-Calculus, \cite{pearl2009causality}]{\label{docalculus}}
%Let $\mathcal{G}$ a CGM and $P_{\mathcal{G}}$ the probability measure induced by the model; then, for disjoint sets of nodes $X,Y,Z,W$ it holds:
%\begin{itemize}
%\item If for the graph $\mathcal{G}_{\bar{X}}$ it holds that $Y$ is conditionally independent from  $Z$ given $X$ and $W$, then
%\[ P_{\mathcal{G}}(Y=y | do(X=x), Z=z, W=w) = P_{\mathcal{G}}(Y=y | do(X=x), W=w). \]
%\item If for the graph $\mathcal{G}_{\bar{X}\underline{Z}}$ it holds that $Y$ is conditionally independent from $Z$ given $X$ and $W$, then
%\[ P(Y=y | do(X=x), do(Z=z), W=w) = P(Y=y | do(X=x), Z = z, W=w). \]
%\item Let $Z(W)$ the set of nodes in  $Z$ that aren't ancestors of any node in  $W$ in the graph $\mathcal{G}_{\bar{X}}$.  If $Y$ is conditionally independent from $Z$ given $X$ and $W$ in the graph $G_{\bar{X}, \bar{Z(W)}}$, then
%\[ P(Y=y | do(X=x), do(Z=z), W=w) = P(Y=y | do(X=x), W=w). \]
%\end{itemize}
%\end{theorem}
%\begin{proof}
%See \cite{pearl2009causality}.
%\end{proof}
%\begin{theorem}[Properties of Do-calculus]{\cite{peters2017elements}}\\
%The following statements hold:
%\begin{itemize}
%\item The Do-Calculus is complete; this is, sufficient for deriving every identifiable interventional distribution \citep{huang2006pearl,shpitser2006identification}.
%\item There exists an algorithm capable of finding all of the identifiable interventions \citep{tian2002,huang2006pearl}.
%\item A necessary and sufficient criteria exists for the identifiability of interventional distributions \citep{shpitser2006identification,huang2006pearl}.
%\end{itemize}
%\end{theorem}
%\begin{proof}
%See \cite{peters2017elements}.
%\end{proof}
%The Do-Calculus rules provide a solution for the identifiability problem:
%\begin{corollary} \cite{pearl2009causality}.\\
%A distribution $q=P(y_1,...,y_k | do(x_1),...,do(x_n))$ is identifiable in a causal graphical model $\mathcal{G}$ if there exists a finite sequence of transformations, where each one of them corresponds to any of the rules of the Do-Calculus, that reduce $q$ to a purely observational expression.
%\end{corollary}

@@

\begin{definition}
We define a Causal Decision Problem (CDP) as a tuple\\ $(\Omega, \mathcal{A}, \mathcal{G},\mathcal{E},\mathcal{C},\succeq)$ where $(\Omega, \mathcal{A}, \mathcal{G},\mathcal{E},\mathcal{C})$ is a Causal Environment and $\succeq$ is a preference relation. 
\end{definition}

@@

\subsection{Causal Reinforcement Learning and Bandit Problems}
It is known that human beings conceive their actions on the world as \textit{intervening} in the world \citep{hagmayer2009decision}. Following this idea, \cite{lattimoreNIPS2016} consider decision problems in which the action to be chosen is an intervention over a causal graphical model in which the conditional probabilities are assumed to be known. The agent must choose the intervention that maximizes the value of a \textit{target variable} after a series of learning rounds. They model their problem by considering that choosing an intervention is choosing an arm of a \textit{slot machine}, usually known as Bandit Problems, in which a gambler chooses an arm and gets some reward. From the rewards they estimate probabilities and output an optimal action in the sense of obtaining minimal regret. The assumption by \cite{lattimoreNIPS2016} of the causal model being known, is later partially relaxed by \cite{sen2017identifying}. In the former, it is mentioned that the case in which the causal model is unknown is left as future work. In our work, we study decision making over unknown causal information.

@@

Consider a Causal Model $G$ and its associated distribution $P_G$ and let $C$ be the set of consequences of interest for a decision maker. Then,
\begin{theorem}[Causal von Neumann-Morgenstern Theorem, \cite{pearl2009causality}]{\label{causal_ut}}
If a rational decision maker faces a Causal Environment and if the causal model is known, then the preference relation $\succeq$ satisfies the von Neumann-Morgenstern rationality axioms if and only if:
\begin{eqnarray*}
&a \succeq b&\\
& \textrm{ if and only if }&\\
 &\sum_{c \in C} P_G(c | do(a))u(c) \geq \sum_{c \in C} P_G(c | do(b))u(c).&
\end{eqnarray*} 
\end{theorem}
Equivalently, the action that must be chosen is 
\begin{equation}
a^\ast = \textrm{argmax}_{a \in \mathcal{A}} \sum_{c \in C} P_G(c | do(a))u(c).
\end{equation}
\begin{proof}
This Theorem assumes the existence of a Causal Graphical Model known by the decision maker; then, by Theorem \ref{vNM}, we know that the action that must be chosen is the one that maximizes the expected utility; we must now ask the expected utility of what? The answer lies on the Causal Graphical Model which controls the relation between actions and consequences; if this is the case, then the probability of a consequence given an action $a$ is in fact $P_G(c|do(a))$. The rest follows by Theorem \ref{vNM}. Details and discussion in Section 4.1 of \cite{pearl2009causality}.
\end{proof}

@@

\section{Future Work}
We have left out the implementation of the algorithm stated in Section \ref{random_graphs} we considered the problem of a decision maker situated in a causal environment who does not fully knows the causal model which controls her environment; previous works have considered different levels of uncertainty over the causal model: knowing conditional distributions, knowing a part of the model, knowing the structure of the model or the variables involved. We leave as future work the case in which the variables of the model are not known and we consider that this scenario could be attacked using tools from Bayesian Nonparametrics \citep{hjort2010bayesian,phadia2013prior} in the form of a prior over causal models. If only a part of the set of variables is known, then a Bayesian Nonparametric prior, which generates causal graphs of distinct sizes and complexity, could capture the patterns of causal interaction between known and possible but unknown variables. This would allow to relax the requirement of causal sufficiency.
